---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Darpan Barua"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<DarpanBarua>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1

library(tidyverse)
library(agricolae)
library(lubridate)
library(here)

#Below loads raw data set 

ntldata <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"))

#Below converts data columns properly
ntldata <- ntldata %>%
  mutate(date = as.Date(paste(year4, daynum), format = "%Y %j"))

#2

#Below sets a consistent theme. 

library(ggplot2)

my_theme <- theme_bw() +
  theme(
    plot.background = element_rect(
      fill = 'linen',
      color = 'Black'
    ),
    legend.background = element_rect(fill = 'linen'),
    axis.line = element_line(
      linewidth = 1,
      color = 'Black'
    ),
    axis.text = element_text(
      family = 'serif'
    ),
    plot.title = element_text(
      
      face = 'bold',
      color = 'Black',
      family = 'serif'
    ),
    panel.background = element_rect(
      fill = "linen",
      colour = "linen",
      linewidth = 0.5,
      linetype = "solid"
    )
  )

# Below sets my_theme as the default theme for all plots
theme_set(my_theme)

```

## Simple regression

Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature during July does not change with depth across all lakes. 
Ha: Mean lake temperature during July changes with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

lake_temp_july <- ntldata %>%
  filter(month(date) == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na()

#5

ggplot(lake_temp_july, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  scale_y_continuous(limits = c(0, 35)) +
  labs(title = "Temperature vs. Depth",
       x = "Depth (m)",
       y = "Temperature (°C)")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The scatter plot suggests that temperature generally decreases as depth increases. While the linear regression line captures the general trend, the spread of points indicates some non-linearity as temperature varies more at shallower depths (~0–5 m) and stabilizes at deeper levels. Deeper water remains consistently cooler, likely due to limited sunlight - making this correlation likely. Therefore, perhaps a non-linear model (e.g., polynomial or logarithmic) might better represent the relationship.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7

lm_model <- lm(temperature_C ~ depth, data = lake_temp_july)
summary(lm_model)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:The linear regression model shows a strong negative relationship between depth and temperature, with 73.87% of the variability in temperature explained by changes in depth (R² = 0.7387). The model is based on 9,726 degrees of freedom, and the effect of depth is highly significant (p < 2.2e-16). The regression equation indicates that for every 1-meter increase in depth, temperature decreases by 1.95°C on average. The intercept (21.96°C) represents the estimated surface temperature, and the residual standard error (3.835°C) suggests some unexplained variability in temperature beyond depth alone.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

#AIC-based model
full_model <- lm(temperature_C ~ year4 + daynum + depth, data = lake_temp_july)
step_model <- step(full_model, direction = "both")
summary(step_model)

#The best model includes depth, day of the year, and year.


#10
#Final model
final_model <- lm(temperature_C ~ depth + year4 + daynum, data = lake_temp_july)
summary(final_model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC-based model selection suggests that the best set of explanatory variables for predicting temperature includes depth, year4 (year), and daynum (day of the year). This final model explains 74.12% of the variance in temperature (R² = 0.7412), indicating a strong fit. When compared to the single-variable model using only depth (R² = 0.7387), this model provides a slight improvement by capturing temporal factors (`year4` and `daynum`). Including `daynum` accounts for within-July seasonal variations, making this a more comprehensive model for purposes of our analysis.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

# ANOVA Model
anova_model <- aov(temperature_C ~ lakename, data = lake_temp_july)
summary(anova_model)

# Linear Model
lm_model <- lm(temperature_C ~ lakename, data = lake_temp_july)
summary(lm_model)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the lakes. According to ANOVA Results- we observe the F-statistic = 50 with a p-value < 2e-16, indicating that at least one lake has a significantly different mean temperature compared to the others. According to Linear Model Results- the intercept (17.66°C) represents the mean temperature of the reference lake. All other lakes have statistically significant coefficients (p < 0.05), meaning their temperatures are significantly different from the reference lake. The R² value (0.0395) is low, which could imply that while lake differences are statistically significant, they explain only a small portion of temperature variability.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

ggplot(lake_temp_july, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(limits = c(0, 35)) +
  labs(title = "Temperature vs. Depth by Lake",
       x = "Depth (m)",
       y = "Temperature (°C)")


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

# Below conducts Tukey's HSD test 

tukey_results <- TukeyHSD(anova_model)
print(tukey_results)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: From the Tukey HSD results, we see that Peter Lake has statistically similar mean temperatures to Paul Lake (p = 0.2242), Ward Lake (p = 0.7827), and West Long Lake (p = 0.0000). No single lake is statistically distinct from all other lakes, but East Long Lake shows significant differences from multiple lakes (p < 0.001 for most comparisons), making it one of the most distinct.


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we were just comparing Peter Lake and Paul Lake, we could use a two-sample t-test which would directly test whether the mean temperatures of Peter Lake and Paul Lake are significantly different, without adjusting for multiple comparisons like Tukey's HSD does! 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

# Below filters data for only Crampton Lake and Ward Lake
lake_subset <- lake_temp_july %>% 
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

# Below conducts the t-test
t_test_result <- t.test(temperature_C ~ lakename, data = lake_subset)

print(t_test_result)


```

>Answer: The two-sample t-test results show a p-value of 0.2649, which is greater than 0.05. This means we fail to reject the null hypothesis, indicating that the mean temperatures of Crampton Lake (15.35°C) and Ward Lake (14.46°C) are not significantly different in July. This result matches what we found in Q16 from the Tukey’s HSD test, where the pairwise comparison between Crampton Lake and Ward Lake had a high p-value (p = 0.9714), suggesting no significant difference in mean temperature.Therefore, both tests confirm that Crampton Lake and Ward Lake have statistically similar mean temperatures. 