---
title: "Assignment 2: Coding Basics"
author: "Darpan Barua"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 55, increasing by fives. Assign this sequence a name.

sequence <- 1 + (0:10)*5


sequence


2.  Compute the mean and median of this sequence.

mean_sequence <- mean(sequence) 


median_sequence <- median(sequence) 


3.  Ask R to determine whether the mean is greater than the median.

mean_greater_than_median <- mean_sequence > median_sequence


mean_greater_than_median



4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. Each integers from 0:10 is multiplied by 5. We add 1 to the product since that's the first number we consider in the counting.  

#2. Here we use the basic functions of mean and median to find the relevant data. We use the variables mean_sequence and median_sequence to denote sepecific mean/median values. 

#3. Here we note a context as "mean_greater_than_median" and denote it as such that mean_sequence is greater than median_sequence. In the end if the condition is true, it should return true; otherwise, false. 

```

## Basics, Part 2

5.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).

student_names <- c("Terry", "Tim", "Nino", "Bethany") #character vectors 


print(student_names) 



test_scores <- c(85, 92, 78, 88 ) #numeric vectors 


print(test_scores)



on_scholarship <- c(TRUE, FALSE, TRUE, FALSE) #logical vectors


print(on_scholarship) 


6.  Label each vector with a comment on what type of vector it is.

#Please check above


7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.


student_data <- data.frame(

    Name = student_names, #Name of the student
    Score = test_scores, #Score obtained by the student on the test 
    Scholarship = on_scholarship #Is the student on scholarship? 
)

print(student_data)


8.  Label the columns of your data frame with informative titles.


#Done. Please check above! 




```{r basics 2}
#5. We create variables and assign the set of components that define those variables. 
#6. We simply set comments next to each line of code. 
#7. We use the data.frame function to create a data frame. 
#8. Comment added next to the data frame elements. 

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Data frame has different types of vector data while a matrix only has one type. 

10. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, print the word "Pass"; otherwise print the word "Fail". 

check_pass_if<-function(score) {

    if (score>50) {

      print("Pass")

  }   else {


      print("Fail")

  }
  
  
}



11. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else `.


check_pass_ifelse<-function(score) {


      print(ifelse(score>50, "Pass", "Fail"))

}


check_pass_ifelse(test_scores)



12. Run both functions using the value 52.5 as the input


check_pass_if(52.5)


check_pass_ifelse(52.5)




13. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

check_pass_if(test_scores) #does not work


check_pass_ifelse(test_scores) #works 



```{r basics 3}
#10. Create a function using if...else - this was a basic function which I denoted using variable "check_pass_if"; here the outcome is distinctive for 'if' test score is 50+ and 'else' if the test score is 50 or 50-. 

#11. Create a function using ifelse() - this was a shorter function which I denoted using variable "check_pass_ifelse"; here the outcome is stated in one line of code, if score is 50+, outcome is "Pass", if not "Fail". 

#12a. Run the first function with the value 52.5 - Inserted 52.5 in place of 'test_scores' in the 'check_pass_if' function 

#12b. Run the second function with the value 52.5  - Inserted 52.5 in place of 'test_scores' in the 'check_pass_ifelse' function 

#13a. Run the first function with the vector of test scores - Inserted 'test_scores' in the 'check_pass_if' function. However, it doesn't work. Reason detailed later in Q14. 

#13b. Run the second function with the vector of test scores - Inserted 'test_scores' in the 'check_pass_ifelse' function. 
```

14. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization")

> Answer:'if'...'else' is designed for singular value inputs (scalar) and it only works with one number at a time. 'ifelse' is designed for entire vectors (vectorized) and therefore applying the condition element-wise to each score in the vector. 


**NOTE** Before knitting, you'll need to comment out the call to the function in Q13 that does not work. (A document can't knit if the code it contains causes an error!)